---
title: "Estimating a Poisson Regression Model with R"
author: "Rémi Piatek"
date: "May 6, 2019"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

## The R Programming Language

- `R`: Popular **open-source programming language** for statistical analysis.
- Widely used in statistics and econometrics.
- **User-friendly and powerful IDE** for `R`: [RStudio](https://www.rstudio.com/).
- Basic functionalities of `R` can be extended by **packages**.
- Large number of packages available on the 
[Comprehensive R Archive Network](https://cran.r-project.org/).
- **Goal of this presentation:** Illustrate how to use `R` for the estimation of a
Poisson regression model.

## Count Data Models

- **Count data** models are used to explain dependent variables that are natural 
numbers, i.e., positive integers such that $y_i \in \mathbb{N}$, where 
$\mathbb{N} = \{0, 1, 2,\ldots\}$. 

- Count data models are frequently used in economics to study **countable events**:
Number of years of education, number of patent applications filed by companies, 
number of doctor visits, number of crimes committed in a given city, etc. 

- The **Poisson model** is a popular count data model.


## Poisson Regression Model

- Given a parameter $\lambda_i > 0$, the **Poisson model** assumes that the 
probability of observing $Y_i=y_i$, where $y_i\in\mathbb{N}$, is equal to:
$$Prob(Y_i = y_i \mid \lambda_i) = \frac{\lambda_i^{y_i}\exp\{-\lambda_i\}}{y_i!},$$
for $i=1,\ldots,N$.

- The mean and the variance of $Y_i$ are equal to the parameter $\lambda_i$:  
$$E(Y_i\mid\lambda_i) = V(Y_i\mid\lambda_i) = \lambda_i,$$
implying *equi-dispersion* of the data.

- To control for **observed characteristics**, the parameter $\lambda_i$ can be 
parametrized as follows (implying $\lambda_i > 0$):
$$E(Y_i|X_i,\beta) \equiv \lambda_i = \exp\{X_i'\beta\},$$
where $X_i$ is a vector containing the covariates.



## Simulating Data

- `R` function simulating data from Poisson regression model:
```{r}
simul_poisson <- function(n, beta) {
  k <- length(beta)               # number of covariates
  x <- replicate(k - 1, rnorm(n)) # simulate covariates
  x <- cbind(1, x)                # for intercept term
  lambda <- exp(x %*% beta)       # individual means
  y <- rpois(n, lambda)           # simulate count
  return(data.frame(y, x))        # return variables
}
```
- Using function to generate data:
```{r}
set.seed(123)
nobs <- 1000
beta <- c(-.5, .4, -.7)
data <- simul_poisson(nobs, beta)
```

## Data Description

- Descriptive statistics:
```{r}
# extract variables of interest from data set
y <- data[, 1]
x <- as.matrix(data[, 2:4])

# descriptive statistics
library(psych)
describe(data)
```

## Data Description

- Histogram of count variable:
```{r, fig.height=5}
barplot(table(y))
```

## Data Description

- Relationship between count variable and covariates:
```{r, fig.height=5}
par(mfrow = c(1, 2))
plot(y, x[, 2])
plot(y, x[, 3])
```

## Likelihood Function and ML Estimator

- Individual contribution to the likelihood function:
$$L_i(\beta;y_i,x_i) = \frac{\exp\{y_ix_i\beta\}\exp\{-\exp\{x_i\beta\}\}}{y_i!}$$
- Individual log-Likelihood function:
$$\ell_i(\beta;y_i,x_i) = \log L_i(\beta;y_i,x_i) 
= y_ix_i\beta - \exp\{x_i\beta\} - \log(y_i!)$$

- Maximum Likelihood Estimator:
$$\hat{\beta}_{\text{MLE}} = \arg\max_{\beta} \sum_{i=1}^N \ell(\beta;y,X)$$

- Optimization (using *minimization* of objective function): 
$$\hat{\beta}_{\text{MLE}} = \arg\min_{\beta} Q(\beta;y,X) \qquad
Q(\beta;y,X) = -\frac{1}{N}\sum_{i=1}^N \ell_i(\beta;y_i,x_i)$$


## Coding the Objective Function

```{r}
# Objective function of Poisson regression model
obj_poisson <- function(beta, y, x) {
  lambda <- x %*% beta
  llik <- y*lambda - exp(lambda) - lfactorial(y)
  return(-mean(llik))
}

# Evaluating objective function
beta0 <- c(1, 2, 3)
obj_poisson(beta0, y, x)
```

## Maximizing the Objective Function

- Set starting values:
```{r}
beta0 <- rep(0, length(beta))
```
- Optimize using quasi-Newton method (BFGS algorithm):
```{r}
opt <- optim(beta0, obj_poisson, method = "BFGS", 
             y = y, x = x)
```
- Show results:
```{r}
cat("ML estimates:", opt$par,
    "\nObjective function:", opt$value, "\n")
```

## Comparing Results to Built-in Function

```{r}
opt_glm <- glm(y ~ 0 + x, family = poisson)
summary(opt_glm)
```

## Comparing Results to Built-in Function

- Collect results from the two approaches to compare them:
```{r}
res <- cbind("True" = beta, "MLE" = opt$par, 
             "GLM" = opt_glm$coefficients)
row.names(res) <- c("constant", "x1", "x2")
res
```

- **Question:** Our results (`MLE`) are virtually the same as those obtained 
with the built-in function `GLM`, but not identical. Where do the small 
differences come from?


## Empirical Illustration

- Goal: Investigate the determinants of fertility.
- Poisson regression model used to estimate the relationship between explanatory
variables and count outcome variable.
- Both our estimator coded from scratch and `R` built-in function will be used.

## Data

- Source: Botswana's 1988 Demographic and Health Survey.
- Data set borrowed from Wooldridge:
```{r}
library(wooldridge)
data(fertil2)
```
- Outcome variable: Total number of living children:
```{r}
y_lab <- "children"
```
- Explanatory variables: Education, age, marital status, living in urban area,
having electricity/TV at home:
```{r}
x_lab <- c("educ", "age", "agesq", "evermarr", "urban", 
           "electric", "tv")
```

## Loading data

- Selecting variables and removing missing values:
```{r}
data <- fertil2[, c(y_lab, x_lab)]
data <- na.omit(data)
```

- Show first 6 observations on first 8 variables:
```{r}
head(data[, 1:8], n = 6)
```

## Descriptive Statitics

```{r}
library(psych)
describe(data)
```

## Plot

```{r, fig.height=5}
attach(data)
par(mfrow = c(1, 2))
blue_transp <- adjustcolor("blue", alpha.f = 0.1)
plot(age, children, pch = 19, col = blue_transp)
plot(educ, children, pch = 19, col = blue_transp)
```


## MLE of the Poisson Model

- Maximum likelihood function using built-in function `glm()`:
```{r}
mle <- glm(children ~ educ + age + agesq + evermarr + 
             urban + electric + tv,
           family = "poisson", data = data)
```
- Maximum likelihood function using our own function:
```{r}
y <- data[, y_lab]
x <- as.matrix(data[, x_lab])
x <- cbind(1, x)          # for intercept term
beta0 <- rep(0, ncol(x))  # starting values
opt <- optim(beta0, obj_poisson, method = "BFGS", 
             y = y, x = x)
```

## MLE of the Poisson Model

- Results different from `glm()`? 
- Optimization algorithms are iterative methods that rely on different criteria
to dertermine if/when the optimum has been reached.
- For example: Change in the objective function, change in the parameter values,
change in the gradient, step size, etc.
- *[More in Advanced Microeconometrics course].*
- Try to adjust tuning parameters, for example add 
`control = list(ndeps = rep(1e-8, ncol(x)))` to `optim()` to change step size 
of gradient approximation.


## Summarizing the Empirical Results

```{r}
summary(mle)
```

## Fitted Values
```{r, fig.height=6}
plot(density(mle$fitted.values), 
     main = "Density of fitted mean values")
```

## Formatting the results
```{r}
library(xtable)
xtable(mle)
```
